---
title: "Exploring the Opioid Epidemic"
author: "Justin Cohler"
date: "January 27, 2019"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE}
library(tidyverse)
library(geojsonio)
library(here)
library(RColorBrewer)
library(cartogram)
library(broom)
library(rgeos)
library(mapproj)
library(dplyr)
library(rmarkdown)
library(tidyverse)
library(here)
library(ggrepel)
library(viridis)
library(lubridate)
library(gghighlight)
library(ggthemes)
library(extrafont)
library(devtools)
library(scales)
library(openintro)
library(gganimate)
library(gifski)
library(gapminder)
library(png)
library(dplyr)
library(rmarkdown)
```

```{r echo=FALSE}
# Pop Per State Data (source: https://www.census.gov/data/tables/2017/demo/popest/nation-total.html)
pop=read.table("https://www.r-graph-gallery.com/wp-content/uploads/2018/01/pop_US.csv", sep=",", header=T)
pop$State <- state2abbr(pop$ï..state)
pop$StateName <- pop$ï..state
```

```{r Plot1}
df_od <- read_csv(here::here("data", "VSRR_Provisional_Drug_Overdose_Death_Counts.csv"))
df_od$Date <- parse_date_time(paste(tolower(df_od$Month), df_od$Year), orders=c("by"))
df_od$NumDeaths <- df_od$`Data Value`
df_od <- df_od %>% filter(`Indicator`=='Number of Deaths' & `State` != 'US')
df_od <- df_od %>% left_join(., pop, by=c("State"="State"))
df_od$FatalitiesPerCap <- df_od$`Data Value`/df_od$pop*100
glimpse(df_od)


p1 <- ggplot(df_od, aes(x=Date, y=`FatalitiesPerCap`, group=`StateName`, colour=`StateName`)) +
  geom_line() +
  gghighlight(max_highlight = 8L, max(`FatalitiesPerCap`),  use_direct_label=FALSE, label_key=`FatalitiesPerCap`) +
  labs(title="West Virginia's Opioid Problem Stands\nIn Stark Contrast to Rest of Country", 
       subtitle = "Drug Overdoses Per Capita From the Previous 5 Years",
       caption="National Center for Health Statistics",
       x="Date", y="% Fatal Drug Overdoses Per Capita") +
  theme(panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.grid.minor.y = element_blank(),
        text=element_text(family="Lato"))

p1
```

Arguably no one has been hit harder by the opioid epidemic than those living in the rust belt and the rural south. One can see above that rural America (Alabama, Arkansas, Kentucky, Tennesee, and West Virginia) comprise some of the states hit hardest per capita by opioid related deaths. Many factors contribute to this scourge on the region, including economic disenfranchisement and high rates of disability enrollment. 

```{r Plot2}
df_ods_by_state <- aggregate(df_od, by=list(df_od$StateName), FUN=mean, na.rm=TRUE)
df_ods_by_state$StateName <- df_ods_by_state$Group.1
df_ods <- df_ods_by_state %>% select(StateName, FatalitiesPerCap)
mean_ods <- mean(df_ods$FatalitiesPerCap)

df_ods$Above <- df_ods$FatalitiesPerCap > mean_ods
glimpse(df_ods)
df_ods$Above[df_ods$Above == TRUE] <- "RED"
df_ods$Above[df_ods$Above == FALSE] <- "GREEN"


p2 <- ggplot(df_ods, aes(x=reorder(StateName, FatalitiesPerCap), y=FatalitiesPerCap, label=format(FatalitiesPerCap, digits=2, nsmall=2))) + 
  geom_point(stat='identity', aes(col=Above), size=6)  +
  scale_color_manual(name="Fatalities By Overdose (% of Pop)", 
                     labels = c("Below Average", "Above Average"), 
                     values = c("GREEN"="#00ba38", "RED"="#f8766d")) +
  geom_text(color="black", size=2) +
  labs(title="New York and West Virginia Lie on Opposite\nPoles of the Opioid Epidemic", 
       subtitle="Fatal Drug Overdoses By State",
       caption="National Center for Health Statistics",
       x="State", y="Fatalities Per Capita (% of Population)") + 
  ylim(0, 2.0) +
  coord_flip() + 
  theme(text=element_text(family="Lato"))
p2
```
While West Virginia is most intensely affected among US states, no state is immune to the epidemic; even in New York, the state whose opioid-related death rate is lowest per capita, one half of one percent of the population will die from a drug-related overdose. The standard deviation of opioid abuse's impact on states is smaller than one might guess if they are only reading articles (link to article) about the plight of West Virginia.

```{r LoadDrugUtilization}
DRUG_NAMES <- c("VICODIN", "OXYCODONE", "OXYCONTIN", "PERCOCET", 
                "OPANA", "KADIAN", "AVINZA", "FENTANYL")

fname_drug_utilization <- "State_Drug_Utilization_Data_"
df_util <- read_csv(here::here("data", paste(fname_drug_utilization, "2008", ".csv", sep=""))) %>% filter(`Product Name` %in% DRUG_NAMES)
for (i in 2009:2018) {
  df_util <- rbind(df_util, read_csv(here::here("data", paste(fname_drug_utilization, i, ".csv", sep=""))) %>% filter(`Product Name` %in% DRUG_NAMES)) 
}
```

```{r Plot3}
df_rx <- aggregate(df_util[,c("Number of Prescriptions", "Total Amount Reimbursed")], by=list(df_util$Year, df_util$`Product Name`), FUN="sum", na.rm=TRUE)
df_rx$`Drug Name` <- df_rx$Group.2


p3 <-  df_rx %>% ggplot(aes(Group.1, `Total Amount Reimbursed`, size=`Number of Prescriptions`, colour=`Drug Name`)) +
  geom_point(alpha = 0.7) +
  geom_line(size=1) + 
  scale_y_continuous(labels = dollar) +
  labs(title = "Pankiller Reimbursements Shrink While\nPrescriptions Remain Fairly Constant",
       subtitle = "Expense Reimbursement & Prescriptions over the past 5 years",
       caption = "Data.Medicaid.gov",
       x = 'Year', y = 'Total $ Reimbursed') +
  theme(panel.background = element_rect(fill = "white", colour = "grey50"),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.grid.minor.y = element_blank(),
        text=element_text(family="Lato"))

p3$labels$size <- "Number of Prescriptions"
p3$labels$colour <- "Drug Name"

p3
```
Above, we use a proxy for presciprt

```{r Plot4}
glimpse(df_util)
#df_util$StateName <- abbr2state(df_util$State)

# Drug Overdoses
DATA_OVERDOSE <- "VSRR_Provisional_Drug_Overdose_Death_Counts.csv"
df_od <- read_csv(here::here("data", DATA_OVERDOSE))
df_od$Date <- parse_date_time(paste(tolower(df_od$Month), df_od$Year), orders=c("by"))
df_od$NumDeaths <- df_od$`Data Value`
df_od <- df_od %>% filter(`Indicator`=='Number of Deaths' & `State` != 'US')
df_od <- df_od %>% left_join(., pop, by=c("State"="State")) %>% left_join(., df_util, by=c("State"="State", "StateName"="StateName", "Year"="Year"))

df_od$FatalitiesPer <- df_od$`Data Value`*1000/df_od$pop
df_od$`Number of Prescriptions` <- df_od$`Number of Prescriptions`*1000/df_od$pop

for (yr in 2008:2018) {
  
  df_ods_by_state <- df_od %>% filter(Year == yr)
  df_ods_by_state <- aggregate(list(FatalitiesPer=df_ods_by_state$FatalitiesPer, NumRxPer=df_ods_by_state$`Number of Prescriptions`), by=list(StateName=df_ods_by_state$StateName), FUN="sum", na.rm=TRUE)

  # Import Cartogram Polygons, merge drug overdose data
  spdf <- geojson_read(here::here("data", "us_states_hexgrid.geojson"),  what = "sp")
  spdf@data = spdf@data %>% mutate(google_name = gsub(" \\(United States\\)", "", google_name)) 
  spdf@data = spdf@data %>% left_join(., df_ods_by_state, by=c("google_name"="StateName"))
  cartogram <- cartogram_cont(spdf, 'FatalitiesPer')
  
  # tidy data to be drawn by ggplot2 (broom library of the tidyverse)
  carto_fortified <- tidy(cartogram, region = "google_name") %>% left_join(. , cartogram@data, by=c("id"="google_name")) 
  centers <- cbind.data.frame(data.frame(gCentroid(cartogram, byid=TRUE), id=cartogram@data$iso3166_2))   

  # plot
  ggplot() +
    geom_polygon(data = carto_fortified, aes(fill = FatalitiesPer, x = long, colour=NumRxPer, y = lat, group = group) , size=0.05, alpha=0.9, color="#f5f5f9") +
    #scale_fill_gradientn(colours=brewer.pal(9, "Greens"), name="Medicaid Opioid Prescriptions", breaks=c(5000, 10000, 25000), guide=guide_legend( keyheight = unit(3, units = "mm"), keywidth=unit(15, units = "mm"), title.position = 'top', label.position = "bottom") ) +
    geom_text(data=centers, aes(x=x, y=y, label=id), color="#444444", size=3, alpha=0.6) +
    theme_void() +
    ggtitle( paste("Where Overdose Deaths Hit Hardest\n",yr) ) +
    theme(
      legend.position = c(0.5, 0.9),
      legend.direction = "horizontal",
      text = element_text(color = "#333333"),
      plot.background = element_rect(fill = "#f5f5f9", color = NA), 
      panel.background = element_rect(fill = "#f5f5f9", color = NA), 
      legend.background = element_rect(fill = "#f5f5f9", color = NA),
      plot.title = element_text(size= 22, hjust=0.5, color = "#444444", margin = margin(b = -0.1, t = 0.4, l = 2, unit = "cm"))
    ) +
    coord_map()  

  ggsave(here::here("plots", paste("cartogram_",yr,".png")))
}

```